\chapter{Conclusions and Future Work}
\epigraph{``I don't see that human intelligence is something that humans can never understand."''}{\vspace{10pt}John McCarthy, March 1989}
\label{chapter:conclusions-fw}

\newpage

This chapter presents the final conclusions obtained through the research effort and the future work that can be derived. 



\section{Conclusions}

This project focused on evaluating the impact of modifying the MLP head in a Visual Transformer-based Facial Expression Recognition model specifically using a challenging or \gls{in-the-wild} based Dataset and using \gls{POSTER} as the baseline architecture without changes in the MLP classification head \cite{zheng_poster_2022}. This allows to have an alternative perspective on improving Facial Expression Recognition performance by focusing on classification head design rather than backbone scaling for \gls{in-the-wild} Facial Expression Recognition (\gls{FER}) images. 

Throughout the work developed in the thesis, a new version of the head is presented as an opportunity to explore how the final classification can help improve model predictability or also achieve better results on challenging classes without modifying the visual transformer's backbone. The new head was integrated into the code  using the same hyperparameters as the baseline and tested with three transformer's sizes (Base/Small/Large).

%
%\begin{itemize}
%	
%	\item  Conclusion 1: The experimental results confirm the research hypothesis that modifying the MLP head leads to improved FER performance compared to the baseline POSTER architecture.
%	
%	\item Conclusion 2.
%	
%	\item Conclusion 3.
%	
%	
%\end{itemize}

\textbf{Conclusion 1 - Head modifications can improve FER performance in a nuanced way.}


The experimental results partially confirm the research hypothesis. Across all six configurations, overall top‑1 accuracy remained stable $(\approx 0.923–0.928)$, indicating that the choice of head architecture does not drastically affect dataset‑level accuracy. This observation is further supported by McNemar’s tests, which reveal no statistically significant differences in paired accuracy outcomes between the baseline and enhanced heads. These results indicate that the proposed head modification does not consistently outperform the baseline at the global level, but also does not introduce additional systematic errors or compromise model generalization.


At the same time, macro‑F1 and per‑class metrics reveal localized and configuration‑dependent changes, particularly in the Small and Base \gls{POSTER} variants. Although bootstrap resampling on macro‑averaged F1 shows that these differences are not statistically significant, the descriptive results suggest that the enhanced head can positively influence prediction quality at a finer granularity, particularly at the class level. Overall, these findings indicate that while global accuracy is largely preserved, head design can modulate class‑level behavior without jeopardizing overall classification power, especially for underrepresented or more ambiguous emotion classes (e.g., Fear, Sad, and Disgust)~\cite{kollias_affect_2021, zhang_leave_2023, wang_survey_2024}.

\textbf{Conclusion 2 – The enhanced head preserves global performance while benefiting difficult emotion classes}

Detailed analysis of class‑wise precision, recall, and F1‑score shows that the enhanced head tends to improve performance for challenging categories such as Fear and Disgust, which are both underrepresented and harder to discriminate. These changes are achieved without degrading, and in some cases slightly improving, well‑defined majority classes such as Happy~\cite{kollias_affect_2021, zhang_leave_2023, wang_survey_2024}. Although bootstrap analysis indicates that the observed macro‑F1 differences are not statistically significant, the consistent direction of class‑wise changes across several configurations suggests that the enhanced head alters the distribution of errors in a controlled manner rather than introducing instability.


This behavior can be attributed to the combination of non‑linear transformations, normalization, and regularization mechanisms in the enhanced head, which allow the model to better exploit subtle facial cues that a purely linear MLP head may struggle to capture. As a result, the enhanced head increases robustness to \gls{intra-class} variability and class imbalance while maintaining overall predictive reliability.


\textbf{Conclusion 3 - Smaller \gls{POSTER} variants with enhanced heads offer an efficient accuracy–cost trade‑off.}

The comparison between Small, Base, and Large \gls{POSTER} models indicates that increasing backbone size does not yield substantial gains in global accuracy or macro‑F1 when the classification head is properly designed. Statistical tests confirm that performance differences between baseline and enhanced heads are not significant, reinforcing the conclusion that smaller models can preserve classification power without sacrificing generalization.

Training time analysis further shows that Small and Base variants are significantly more efficient while achieving performance comparable to the Large model. Consequently, from a practical perspective, deploying and training a Small or Base \gls{POSTER} model with the enhanced head represents a favorable trade‑off for \gls{in-the-wild} \gls{FER} applications, offering competitive benchmark performance with reduced computational and resource requirements.


\section{Future Work}

The current investigation allows to have a lot opportunities to outperform the \gls{POSTER} \cite{zheng_poster_2022} research head investigation 


\begin{itemize}
\item A usage of another dataset for validation is recommended to allow review head changes performance on the proposed head. This permits to have a more valid experiment and metrics for a proper benchmarking comparisson. The dataset Affecnet\footnote{\url{https://www.mohammadmahoor.com/pages/databases/affectnet/}}\footnote{\url{https://www.kaggle.com/datasets/fatihkgg/affectnet-yolo-format}} is used on the baseline architecture paper ~\cite{zheng_poster_2022}. 

Although RAF-DB was selected as the primary dataset for the thesis work due to its balanced annotations and widespread adoption in transformer-based \gls{FER} researches and state of the art papers, other commonly used datasets include FER2013 \footnote{\url{https://www.kaggle.com/datasets/msambare/fer2013}}~\cite{khaireddin_facial_2021} and AffectNet~\cite{mollahosseini_affectnet_2019}, which are often employed for large-scale or cross-dataset generalization studies.

\item  The enhanced head architecture explored in this work represents only one point in a broader design space. Future work could investigate deeper or multi‑branch heads, gated or adaptive activation functions (e.g., GLU‑based layers), residual MLP blocks, or attention‑augmented classification heads. Systematic ablation studies comparing these variants would help isolate which components (normalization, dropout, depth, activation type) most strongly contribute to performance gains on difficult emotion classes.

\item Whereas this thesis work relied on standard evaluation metrics including accuracy, macro‑averaged F1‑Score, precision, recall, confusion matrices, learning curves and non parametric tests like McNemar and Bootstrap  analysis  additional statistical validation techniques can also explored like for example Cohen’s kappa for agreement analysis, and Chi‑Square tests on confusion matrices. The use of such methods provides richer information about the significance and reliability of observed differences. A valuable line of future work is to formalize and integrate these statistical tests into the evaluation pipeline encouraging more statistically rigorous benchmarking practices in \gls{FER}.
\end{itemize}. 



%\begin{itemize}
%	\item Future work 1.
%	
%	\item Future work 2.
%	
%	\item Future work 3.
%	
%\end{itemize}
